Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-06 tex/main.tex
Minor changes throughout.

\begin{abstract}
SMT solvers have recently been extended with finite model finding techniques.
However, these generally cannot cope with universal quantifiers ranging over
infinite types, such as integers and recursive datatypes. This paper introduces
a translation that eliminates infinite quantifiers in axioms specifying a large
class of recursive functions, including terminating, tail-recursive, and
productive functions. An empirical evaluation confirms that the approach works
well on benchmarks from two sources. The translation is implemented as a
preprocessor in the solver CVC4.
\end{abstract}

@@

For example, the set $\{ \forallf{\con{f}} \vec x.\;\, \con{f}( x ) \teq 0 \}$ is admissible.
An admissible set can include definitions of nonterminating functions, e.g.
$\{ \forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x - 1 ) \}$ or even $\{ \forallf{\con{f}} x.\allowbreak\;\, \con{f}( x ) \teq \con{f}( x ) \}$ are admissible.
On the other hand, the set $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$ is not admissible,
since the empty set of formulas is closed under function expansion with respect to this set,
and there is no model satisfying $\forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1$.
Consider the set $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ), \forallf{\con{g}}.\;\, \lite\bigl( \neg \con{f}( x ) \teq 5, \con{g}( x ) \teq \con{g}( x ) + 1, \con{g}( x ) \teq \con{f}( x ) \bigr) \}$.
While this set has a model where $f$ and $g$ are interpreted as the constant $5$, it is not admissible
since the formula $\con{f}( 0 ) \teq 4$ is closed under function expansion with respect to this set,
but there exists no interpretation satisfying both $\con{f}( 0 ) \teq 4$ and $\forallf{\con{g}}.\;\, \lite( \neg \con{f}( x ) \teq 5, \con{g}( x ) \teq \con{g}( x ) + 1, \con{g}( x ) \teq \con{f}( x ) )$.

@@

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable if and only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{f}$ can be
interpreted as $\typeint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if
and only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\typeint}.\;\, \lite\bigl( x \leq 0,\allowbreak\; \con{s}(x)
\teq 0,\; \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissble.
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargtype{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. The nested existential
$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
a similar constraint is enforced recursively: any model satisfying $\varphi_{\con{s}}[t/x]$
must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$ \ajr{revise this sentence, not quite accurate}. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$, it is easy to see that
due to the structure of formula~(\ref{eq:ex-after}),
$\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}

@@

For solvers, we considered the SMT solver \ziii~\cite{de-moura-bjoerner-2008},
which runs both heuristic methods for quantifier instantiation~\cite{DBLP:conf/cade/MouraB07},
as well as methods for finding models for quantified formulas~\cite{GeDeM-CAV-09}.
We also considered three configurations of \cvc~\cite{barrett-et-al-2011} which
we will refer to as \cvcd, \cvcf, and \cvci.
The default configuration \cvcd runs heuristic quantifier instantiation,
but does not include techniques for finding models.
The configuration \cvcf runs heuristic instantiation and the finite model
finding procedure described in~\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}. The
last configuration \cvci incorporates techniques for automating inductive
reasoning in SMT~\cite{reynolds-kuncak-2015}.
%All configurations of \cvc incorporate techniques for conflict-based instantiation~\cite{} ..?

@@

The translation $\conv$ significantly improves the effectiveness of known SMT
techniques for finding counterexamples for formulas involving recursive
function definitions.
For the convenience of end users, it
has been implemented in the latest development version of \cvc %~\cite{ReyEtAl-1-RR-13}
(version~1.5 prerelease).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written in using the $\definefunreccmd$ command,
from the SMT-LIB 2.5 draft standard \cite{smtlib25}.
For example, formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as
%
\begin{quote}
\begin{verbatim}
(define-fun-rec s ((n Int)) Int (ite (<= n 0) 0 (+ n (s (- n 1)))))
(declare-fun c () Int)
(assert (> (s c) 100))
(check-sat)
\end{verbatim}
\end{quote}
%
When reading this input,
\cvc adds the annotated quantified formula
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
If \cvc's finite model finding mode for recursive functions is enabled (using
the command-line option \texttt{--fmf-fun}), it will replace its list of known
assertions based on the conversion $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it
used for recursive function definitions. For our example, it will output the following
model:
\ajr{prefer having the full output like this here?}
\jb{I like it the way you did it now, but I took the freedom to reformat the output slightly
(to avoid the small font---I generally try to avoid ad hoc font mixtures---and to show
the nesting more clearly).}
%it may output the interpretation~(\ref{eq:approx-interp}) from Section~\ref{sec:encoding} for $s$.
%
\begin{quote}
\begin{verbatim}
(model
  (define-fun s (($x1 Int)) Int 
     (ite (= $x1 14) 105 (ite (= $x1 13) 91 (ite (= $x1 12) 78 
        (ite (= $x1 11) 66 (ite (= $x1 10) 55 (ite (= $x1 4) 10 
           (ite (= $x1 9) 45 (ite (= $x1 8) 36 (ite (= $x1 7) 28 
              (ite (= $x1 6) 21 (ite (= $x1 3) 6 (ite (= $x1 5) 15 
                 (ite (= $x1 2) 3 (ite (= $x1 1) 1 0)))))))))))))))
  (define-fun c () Int 14)
)
\end{verbatim}
\end{quote}

The \texttt{--fmf-fun} option instructs \cvc to assume that the
functions provided by the user are admissible.
Admissibility must be checked separately---for example, using a
syntactic criterion or a termination prover.
If this criterion is not met, the solver may answer ``satisfiable'' on
an unsatisfiable problem. For example, if we add the inconsistent definition
%
\begin{quote}
\begin{verbatim}
(define-fun-rec e ((n Int)) Int (+ (e n) 1))
\end{verbatim}
\end{quote}
%
to the above problem, \cvc will still find the same ``model''
if \texttt{--fmf-fun} is specified.

@@

We described the most closely related work, by Ge and de Moura
\cite{GeDeM-CAV-09} and by Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}, in the introduction already.
Outside the world of SMT, model finding has been
studied many times. The text below is partly adapted from Blanchette's Ph.D.\
thesis \cite{blanchette-2012-phd}.

>>>>

\begin{abstract}
SMT solvers have recently been extended with techniques for finding models
in presence of quantified formulas.
However, these generally cannot cope with universal quantifiers ranging over
infinite types, such as integers and recursive datatypes. This paper introduces
a translation that eliminates infinite quantifiers in axioms specifying a large
class of recursive functions, including terminating, tail-recursive, and
productive functions. An empirical evaluation confirms that the approach works
well on benchmarks from two sources. The translation is implemented as a
preprocessor in the solver CVC4.
\end{abstract}

@@

For example, the set $\{ \forallf{\con{f}} \vec x.\;\, \con{f}( x ) \teq 0 \}$ is admissible.
An admissible set can include definitions of nonterminating functions, e.g.
$\{ \forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x - 1 ) \}$ or even $\{ \forallf{\con{f}} x.\allowbreak\;\, \con{f}( x ) \teq \con{f}( x ) \}$ are admissible.
On the other hand, the set $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$ is not admissible,
since the empty set of formulas is closed under function expansion with respect to this set,
and there is no model satisfying $\forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1$.
Consider the set $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ), \forallf{\con{g}} x.\;\, \lite\bigl( \neg \con{f}( x ) \teq 5, \con{g}( x ) \teq \con{g}( x ) + 1, \con{g}( x ) \teq \con{f}( x ) \bigr) \}$.
While this set has a model where $f$ and $g$ are interpreted as the constant $5$, it is not admissible
since the formula $\con{f}( 0 ) \teq 4$ is closed under function expansion with respect to this set,
but there exists no interpretation satisfying both $\con{f}( 0 ) \teq 4$ and $\forallf{\con{g}} x.\;\, \lite( \neg \con{f}( x ) \teq 5, \con{g}( x ) \teq \con{g}( x ) + 1, \con{g}( x ) \teq \con{f}( x ) )$.

@@

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable if and only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{f}$ can be
interpreted as $\typeint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if
and only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\typeint}.\;\, \lite\bigl( x \leq 0,\allowbreak\; \con{s}(x)
\teq 0,\; \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissble.
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargtype{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. The nested existential
$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
a similar constraint is enforced recursively: when $\varphi_{\con{s}}[t/x]$ is used in part to evaluate $s( c )$,
any model of $\varphi_{\con{s}}[t/x]$ must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$, it is easy to see that
due to the structure of formula~(\ref{eq:ex-after}),
$\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}

@@

For solvers, we considered the SMT solver \ziii~\cite{de-moura-bjoerner-2008},
which runs both heuristic methods for quantifier instantiation~\cite{DBLP:conf/cade/MouraB07},
as well as methods for finding models for quantified formulas~\cite{GeDeM-CAV-09}.
We also considered three configurations of \cvc~\cite{barrett-et-al-2011} which
we will refer to as \cvcd, \cvcf, and \cvci.
The default configuration \cvcd runs heuristic and conflict-based techniques for quantifier instantiation~\cite{ReynoldsTinelliMoura14},
but does not include techniques for finding models.
The configuration \cvcf runs heuristic instantiation and the finite model
finding procedure described in~\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}. The
last configuration \cvci incorporates techniques for automating inductive
reasoning in SMT~\cite{reynolds-kuncak-2015}.

@@

The translation $\conv$ significantly improves the effectiveness of known SMT
techniques for finding counterexamples for formulas involving recursive
function definitions.
For the convenience of end users, it
has been implemented in the latest development version of \cvc %~\cite{ReyEtAl-1-RR-13}
(version~1.5 prerelease).
Function definitions $\forallf{\con{f}} \vec x.\; \varphi$ can be written in using the $\definefunreccmd$ command,
from the SMT-LIB 2.5 draft standard \cite{smtlib25}.
For example, formula~(\ref{eq:ex-before}) from Example~\ref{ex:translation}
can be specified as
%
\begin{quote}
\begin{verbatim}
(define-fun-rec s ((n Int)) Int (ite (<= n 0) 0 (+ n (s (- n 1)))))
(declare-fun c () Int)
(assert (> (s c) 100))
(check-sat)
\end{verbatim}
\end{quote}
%
When reading this input,
\cvc adds the annotated quantified formula
$$\forallf{\con{s}} x.\;\, \con{s}( x ) \teq \lite\bigl( x \leq 0,\; 0,\; \con{s}( x-1 )\bigr)$$
to its list of assertions,
which after rewriting becomes
$$\forallf{\con{s}} x.\;\, \lite\bigl( x \leq 0,\; \con{s}( x ) \teq 0,\; \con{s}( x ) \teq \con{s}( x-1 ) \bigr)$$
If \cvc's finite model finding mode for recursive functions is enabled (using
the command-line option \texttt{--fmf-fun}), it will replace its list of known
assertions based on the conversion $\conv$ before checking for satisfiability.
%If the functions provided in the input are admissible,
%then a ``satisfiable" response from the solver implies that a model exists for the original formula.
Accordingly, the solver will output the approximation of the interpretation it
used for recursive function definitions. For our example, it will produce the following output:
%\ajr{prefer having the full output like this here?}
%\jb{I like it the way you did it now, but I took the freedom to reformat the output slightly
%(to avoid the small font---I generally try to avoid ad hoc font mixtures---and to show
%the nesting more clearly).}
%it may output the interpretation~(\ref{eq:approx-interp}) from Section~\ref{sec:encoding} for $s$.
%
\begin{quote}
\begin{verbatim}
(model
  (define-fun s (($x1 Int)) Int 
     (ite (= $x1 14) 105 (ite (= $x1 13) 91 (ite (= $x1 12) 78 
        (ite (= $x1 11) 66 (ite (= $x1 10) 55 (ite (= $x1 4) 10 
           (ite (= $x1 9) 45 (ite (= $x1 8) 36 (ite (= $x1 7) 28 
              (ite (= $x1 6) 21 (ite (= $x1 3) 6 (ite (= $x1 5) 15 
                 (ite (= $x1 2) 3 (ite (= $x1 1) 1 0)))))))))))))))
  (define-fun c () Int 14)
)
\end{verbatim}
\end{quote}

The \texttt{--fmf-fun} option instructs \cvc to assume that the
functions provided by the user are admissible.
Admissibility must be checked separately---for example, using a
syntactic criterion or a termination prover.
If this criterion is not met, the solver may answer ``satisfiable'' on
an unsatisfiable problem. For example, if we add the inconsistent definition
%
\begin{quote}
\begin{verbatim}
(define-fun-rec e ((n Int)) Int (+ (e n) 1))
\end{verbatim}
\end{quote}
%
to the above problem, \cvc will still produce the same output
if \texttt{--fmf-fun} is specified.

@@

We described the most closely related work, by Ge and de Moura
\cite{GeDeM-CAV-09} and by Reynolds et al.\
\cite{ReyEtAl-1-RR-13,reynolds-et-al-2013}, in the text already.
Outside the world of SMT, model finding has been
studied many times. The text below is partly adapted from Blanchette's Ph.D.\
thesis \cite{blanchette-2012-phd}.
